Type Error Telemetry at Scale

### 2023-09-28 programming23/paper.tex
quick error-by-mode bars

Each telemetry record contains seven kinds of data.
Refer to~\cref{f:tdata} for an illustration:
\begin{enumerate}
  \item
    ID: a \panon{} identifier for the current session
  \item
    Time: one timestamp from the client and one from the server.
    %% timestamp explanation \cref{s:data-cleaning}
  \item
    Mode: type analysis mode of the current file, which is either \mnocheck{},
    \mnonstrict{}, or \mstrict{}.
  \item
    Reason: a flag that explains whether this record was sent due to a
    randomly-selected keystroke or a module switch.
  \item
    Size Info: number of lines in the codebase and number of lines in the edit range.
  \item
    Overall Counts: summarizes type errors and forced strict errors during the
    last two invocations of type analysis, and additionally reports the number
    of times type analysis hit an internal limit on problem
    size~(see~\cref{s:code-too-complex} for details).
    For each invocation and each kind of
    analysis error, there are three summary counts:
    %% grand total = 12
    \begin{itemize}
      \item total number of errors,
      \item errors in the current module, and
      \item errors in the current edit range.
    \end{itemize}
  \item
    Edit Range Counts: these list specific errors that arose in the last two
    type analyses and overlap with the current edit range.
    Since there are 35 possible errors~(\cref{s:context}), there can be up to 70 counts
    in a telemetry record.
    Exactly how to interpret this data depends on which analysis it came from:
    \begin{itemize}
      \item
        Errors in the latest type analysis are currently visible to the
        creator.
        These may have been introduced by the changes in the edit range.
      \item
        Errors from the previous type analysis were visible before the creator
        made the latest edits.
        These may have motivated some of the changes in the edit range.
    \end{itemize}
\end{enumerate}

@@

\begin{figure}[t]
  \begin{tabular}[t]{ll} \\
    \begin{tabular}[t]{r@{~~}r@{~}l}
      595,137 & type errors \\
      289,698 & in current module & [\pct{48.68}] \\
       30,924 & in edit range & [\pct{5.20}]
    \end{tabular}
    \begin{tabular}[t]{r@{~~}r@{~}l}
      72,235,735 & {forced strict errors} \\
      37,027,281 & in current module & [\pct{51.26}] \\
       1,111,178 & in edit range & [\pct{1.54}]
    \end{tabular}
  \end{tabular}
  \caption{Type errors and forced strict errors across all telemetry records.}
  \label{f:count-analysis-errors}
\end{figure}

@@

\Cref{f:count-analysis-errors} lists the number of type analysis errors
and breaks them down by location in the codebase.
In total, there are nearly 600K type errors.
These appeared in modules that opted in to typechecking with either the
\mnonstrict{} or \mstrict{} mode.
The number of forced strict errors is much larger, 72M, because this analysis
ran on every module using strict type.
(Furthermore, forced strict runs silently, so creators have no
awareness of or incentive to
fix these errors.)
Approximately half the analysis errors occurred in the current module,
and a small fraction overlapped with the current edit range~(\pct{5} for
type errors, \pct{1} for forced strict).

@@

\begin{figure}[t]
  %% strict
  %%  te total 22771
  %%  mean 6.9 std 26.24592591869705 med 1 max 385
  %% 
  %%  fs total 114259
  %%  mean 34.9 std 155.41955823164017 med 1 max 4789
  %% 
  %% nonstrict
  %%  te total 362964
  %%  mean 2.3 std 66.44456626891912 med 0 max 13511
  %% 
  %%  fs total 7546175
  %%  mean 48.6 std 270.60578740369266 med 3 max 16712
  %%
  %% nocheck
  %%  te total 176309
  %%  mean 0.13 std 1.9040337635815197 med 0 max 289
  %% 
  %%  fs total 63784284
  %%  mean 47.7 std 885.8225471699477 med 1 max 251131

  \begin{tabular}{l}
    600K type errors         = \pct{35} \mnocheck{}, \pct{61} \mnonstrict{}, \pct{4} \mstrict{}
    \\
    72M forced scrict errors = \pct{89} \mnocheck{}, \pct{10} \mnonstrict{}, \pct{1} \mstrict{}
    %% really 0.16 strict, very low!!!
  \end{tabular}
  \caption{FILL}
  \label{f:error-by-mode}
\end{figure}

\Cref{f:error-by-mode} shows how the type errors and forced strict errors
in the data are split across type analysis modes.

@@

Forced strict analysis applies the same typechecker to every module, regardless
of mode, in the hope of enabling IDE tools.
Consequently, the breakdown of forced strict errors matches the overall usage
data~(\cref{f:dataset-overview}): almost \pct{90} for \mnocheck{},
\pct{10} for \mnonstrict{}, and \pct{1} for \mstrict{}.
Type analysis finds a large number of errors in \mnocheck{} code;
unfortunately, we have no insight into whether those errors are true
positives or whether the analysis led to useful IDE suggestions.

@@

\begin{figure}[t]\centering
  %% TODO
  %% - thin bars
  %% - colors for good , bad (disagree) , neutral
  \begin{tabular}{lll}
    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
  \end{tabular}
  \caption{Comparing the change in type errors to the change in forced strict errors.}
  \label{f:tefs-compass}
\end{figure}

Since Roblox Studio runs each codebase through two passes of type analysis,
it is interesting to compare their results.
Of special interest is the \mstrict{} mode case, because the number of
type errors and forced strict errors should be nearly the same.
But, there are two differences between \mstrict{} type checking
and forced strict checking:
\begin{itemize}
  \item
    Forced strict checks all dependencies in \mstrict{} mode, no matter
    what mode they declare.
    So, the number of forced strict errors may be higher.
  \item
    Forced strict has looser rules for the \emph{data model};
    that is, for the static assets that a program manipulates.
    So, the number of forced strict errors may be lower.
    %% initial DM, all properties go into a table type; DM -> folder -> model -> part
    %% get dot-driven autocomplete
\end{itemize}
Our data contains a surprising number of records where forced strict
reports fewer errors.
Data model edits are apparently a regular occurrence in Luau programs.

\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in forced
strict errors (\fskey{}) for the three analysis modes.
To a first approximation, the changes either agree with one another
or disagree.
For example, if type errors increase but forced strict errors decrease,
we have a disagreement.
The plots futher categorize agreements and disagreements into
the following five cases:
\begin{enumerate}
  \item
    \emph{zero}: neither \tekey{} nor \fskey{} changed
  \item
    \emph{te only}: \tekey{} changed but \fskey{} remained the same
  \item 
    \emph{fs only}: \tekey{} remained the same but \fskey{} changed
  \item
    \emph{dis}: \tekey{} and \fskey{} changed in opposite directions
  \item 
    \emph{agree}: \tekey{} and \fskey{} changed in the same direction
\end{enumerate}
Both \emph{zero} and \emph{agree} are agreements.
The others are disagreements.
The plots show what percent of all telemetry records fell into these five categories.

@@

Another insightful visualization that the telemetry supports is plots of individual sessions
over time.
\Cref{f:indy-session} shows the type errors, forced strict errors, and module switches
in two \mstrict{} mode sessions.
The x-axis shows time in seconds from the start of the session.
The y-axis counts errors.
Type errors appear as orange plus marks.
Forced strict errors appear as green x marks.
The vertical lines show module switches.

@@

A general observation is that the number of type errors is always less than the number of forced strict
errors, though the numbers are close.
This should happen in \mstrict{} mode.

>>>>

Each telemetry record contains seven kinds of data.
Refer to~\cref{f:tdata} for an illustration:
\begin{enumerate}
  \item
    ID: a \panon{} identifier for the current session
  \item
    Time: one timestamp from the client and one from the server.
    %% timestamp explanation \cref{s:data-cleaning}
  \item
    Mode: type analysis mode of the current file, which is either \mnocheck{},
    \mnonstrict{}, or \mstrict{}.
  \item
    Reason: a flag that explains whether this record was sent due to a
    randomly-selected keystroke or a module switch.
  \item
    Size Info: number of lines in the codebase and number of lines in the edit range.
  \item
    Overall Counts: summarizes type errors and \FS{} errors during the
    last two invocations of type analysis, and additionally reports the number
    of times type analysis hit an internal limit on problem
    size~(see~\cref{s:code-too-complex} for details).
    For each invocation and each kind of
    analysis error, there are three summary counts:
    %% grand total = 12
    \begin{itemize}
      \item total number of errors,
      \item errors in the current module, and
      \item errors in the current edit range.
    \end{itemize}
  \item
    Edit Range Counts: these list specific errors that arose in the last two
    type analyses and overlap with the current edit range.
    Since there are 35 possible errors~(\cref{s:context}), there can be up to 70 counts
    in a telemetry record.
    Exactly how to interpret this data depends on which analysis it came from:
    \begin{itemize}
      \item
        Errors in the latest type analysis are currently visible to the
        creator.
        These may have been introduced by the changes in the edit range.
      \item
        Errors from the previous type analysis were visible before the creator
        made the latest edits.
        These may have motivated some of the changes in the edit range.
    \end{itemize}
\end{enumerate}

@@

\begin{figure}[t]
  \begin{tabular}[t]{ll} \\
    \begin{tabular}[t]{r@{~~}r@{~}l}
      595,137 & type errors \\
      289,698 & in current module & [\pct{48.68}] \\
       30,924 & in edit range & [\pct{5.20}]
    \end{tabular}
    \begin{tabular}[t]{r@{~~}r@{~}l}
      72,235,735 & {\FS{} errors} \\
      37,027,281 & in current module & [\pct{51.26}] \\
       1,111,178 & in edit range & [\pct{1.54}]
    \end{tabular}
  \end{tabular}
  \caption{Type errors and \FS{} errors across all telemetry records.}
  \label{f:count-analysis-errors}
\end{figure}

@@

\Cref{f:count-analysis-errors} lists the number of type analysis errors
and breaks them down by location in the codebase.
In total, there are over 590,000 type errors.
%% delete? upcoming figure has the same info
% Most of these (\pct{65}) appeared in modules that opted in to typechecking with either the
% \mnonstrict{} or \mstrict{} mode; the remainder are syntax errors reported
% by \mnocheck{} mode.
The number of \FS{} errors is much larger, 72M, because this analysis
ran on every module using strict type.
(Furthermore, \FS{} runs silently, so creators have no
awareness of or incentive to
fix these errors.)
Approximately half the analysis errors occurred in the current module,
and a small fraction overlapped with the current edit range~(\pct{5} for
type errors, \pct{1} for \FS{}).

@@

\begin{figure}[t]\centering

  %%%% from nsa aggregate-tefs
  %% strict
  %%  te total 22771
  %%  mean 6.9 std 26.24592591869705 med 1 max 385
  %% 
  %%  fs total 114259
  %%  mean 34.9 std 155.41955823164017 med 1 max 4789
  %% 
  %% nonstrict
  %%  te total 362964
  %%  mean 2.3 std 66.44456626891912 med 0 max 13511
  %% 
  %%  fs total 7546175
  %%  mean 48.6 std 270.60578740369266 med 3 max 16712
  %%
  %% nocheck
  %%  te total 176309
  %%  mean 0.13 std 1.9040337635815197 med 0 max 289
  %% 
  %%  fs total 63784284
  %%  mean 47.7 std 885.8225471699477 med 1 max 251131

  \newcommand{\labelbars}[1]{\begin{tabular}[t]{l@{}l} \begin{tabular}[t]{r} \mnocheck{} \\ \mnonstrict{} \\ \mstrict{} \\ ~ \end{tabular} & #1 \end{tabular}}

    \begin{tabular}[t]{cc}
    %% TODO why text blurry??!
    Type errors & Background errors \\
    \labelbars{\includegraphics[valign=M]{img/error-by-mode-te.pdf}}
    & \labelbars{\includegraphics[valign=M]{img/error-by-mode-fs.pdf}}
  \end{tabular}
  \caption{FILL}
  \label{f:error-by-mode}
\end{figure}

\Cref{f:error-by-mode} shows how the type errors and \FS{} errors
in the data are split across type analysis modes.

@@

Background analysis applies the same typechecker to every module, regardless
of mode, in the hope of enabling IDE tools.
Consequently, the breakdown of \FS{} errors matches the overall usage
data~(\cref{f:dataset-overview}): almost \pct{90} for \mnocheck{},
\pct{10} for \mnonstrict{}, and \pct{1} for \mstrict{}.
Type analysis finds a large number of errors in \mnocheck{} code;
unfortunately, we have no insight into whether those errors are true
positives or whether the analysis led to useful IDE suggestions.

@@

\begin{figure}[t]\centering
  %% TODO
  %% - thin bars
  %% - colors for good , bad (disagree) , neutral
  \begin{tabular}{lll}
    \mnocheck{} & \mnonstrict{} & \mstrict{} \\
    \includegraphics[width=0.3\columnwidth]{img/compass-nocheck.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-nonstrict.pdf}
    &
    \includegraphics[width=0.3\columnwidth]{img/compass-strict.pdf}
  \end{tabular}
  \caption{Comparing the change in type errors to the change in \FS{} errors.}
  \label{f:tefs-compass}
\end{figure}

Since Roblox Studio runs each codebase through two passes of type analysis,
it is interesting to compare their results.
Of special interest is the \mstrict{} mode case, because the number of
type errors and \FS{} errors should be nearly the same.
But, there are two differences between \mstrict{} type checking
and \FS{} checking:
\begin{itemize}
  \item
    Background checks all dependencies in \mstrict{} mode, no matter
    what mode they declare.
    So, the number of \FS{} errors may be higher.
  \item
    Background checking has looser rules for the \emph{data model};
    that is, for the static assets that a program manipulates.
    So, the number of \FS{} errors may be lower.
    %% initial DM, all properties go into a table type; DM -> folder -> model -> part
    %% get dot-driven autocomplete
\end{itemize}
Our data contains a surprising number of records where \FS{}
reports fewer errors.
Data model edits are apparently a regular occurrence in Luau programs.

\Cref{f:tefs-compass} compares changes in type errors (\tekey{}) to changes in
\FS{} errors (\fskey{}) for the three analysis modes.
To a first approximation, the changes either agree with one another
or disagree.
For example, if type errors increase but \FS{} errors decrease,
we have a disagreement.
The plots futher categorize agreements and disagreements into
the following five cases:
\begin{enumerate}
  \item
    \emph{zero}: neither \tekey{} nor \fskey{} changed
  \item
    \emph{te only}: \tekey{} changed but \fskey{} remained the same
  \item 
    \emph{fs only}: \tekey{} remained the same but \fskey{} changed
  \item
    \emph{dis}: \tekey{} and \fskey{} changed in opposite directions
  \item 
    \emph{agree}: \tekey{} and \fskey{} changed in the same direction
\end{enumerate}
Both \emph{zero} and \emph{agree} are agreements.
The others are disagreements.
The plots show what percent of all telemetry records fell into these five categories.

@@

Another insightful visualization that the telemetry supports is plots of individual sessions
over time.
\Cref{f:indy-session} shows the type errors, \FS{} errors, and module switches
in two \mstrict{} mode sessions.
The x-axis shows time in seconds from the start of the session.
The y-axis counts errors.
Type errors appear as orange plus marks.
Background errors appear as green x marks.
The vertical lines show module switches.

@@

A general observation is that the number of type errors is always less than the number of \FS{}
errors, though the numbers are close.
This should happen in \mstrict{} mode.
